---
title: "Assignment 6: Data Scraping"
author: "Gina Givens"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)

getwd()


```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
the_website <- read_html(
  'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024'
)
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
tables <- the_website %>% 
  html_nodes('table')

the_system_name <- the_website %>% 
  html_nodes('td:nth-child(2)') %>% 
  html_text() %>% 
  .[1]

the_pwsid <- the_website %>% 
  html_nodes('td:nth-child(5)') %>% 
  html_text() %>% 
  .[1]

the_ownership <- the_website %>% 
  html_nodes('td:nth-child(4)') %>% 
  html_text() %>% 
  .[2]

the_withdrawal_table <- tables[[13]]

the_withdrawals <- the_withdrawal_table %>% 
  html_nodes('td') %>% 
  html_text() %>% 
  .[c(2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24)]

the_system_name
the_pwsid
the_ownership
the_withdrawals

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
the_month <- c(
  "Jan", "May", "Sep", "Feb", "Jun", "Oct",
  "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"
)

the_df <- data.frame(
  system_name = rep(the_system_name, 12),
  PWSID = rep(the_pwsid, 12),
  ownership = rep(the_ownership, 12),
  max_day_use = the_withdrawals,
  month = the_month
)

the_df <- the_df %>% 
  mutate(
    max_day_use = as.numeric(max_day_use),
    Date = as.Date(paste0(
      "2024-",
      match(month, month.abb),
      "-01"
    ))
  )

the_df
#5 
the_df %>% 
  ggplot(aes(x = Date, y = max_day_use)) +
  geom_line() +
  labs(
    x = "Month",
    y = "Maximum Daily Withdrawal (MGD)",
    title = "Durham Maximum Daily Water Withdrawal, 2024"
  )

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs - "PWSID" and "year" - that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. 
scrape.it <- function(PWSID, year){
  
  # Create the URL
  the_url <- paste0(
    "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",
    PWSID,
    "&year=",
    year
  )
  
  # Read the website
  the_website <- read_html(the_url)
  
  # Scrape the data
  tables <- the_website %>% 
    html_nodes('table')
  
  the_system_name <- the_website %>% 
    html_nodes('td:nth-child(2)') %>% 
    html_text() %>% 
    .[1]
  
  the_ownership <- the_website %>% 
    html_nodes('td:nth-child(4)') %>% 
    html_text() %>% 
    .[2]
  
  the_withdrawal_table <- tables[[13]]
  
  the_withdrawals <- the_withdrawal_table %>% 
    html_nodes('td') %>% 
    html_text() %>% 
    .[c(2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22, 24)]
  
  # Create the dataframe
  the_month <- c(
    "Jan", "May", "Sep", "Feb", "Jun", "Oct",
    "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"
  )
  
  the_df <- data.frame(
    system_name = rep(the_system_name, 12),
    PWSID = rep(PWSID, 12),
    ownership = rep(the_ownership, 12),
    max_day_use = the_withdrawals,
    month = the_month,
    year = rep(year, 12)
  ) %>% 
    mutate(
      max_day_use = as.numeric(max_day_use),
      Date = as.Date(paste0(
        year,
        "-",
        match(month, month.abb),
        "-01"
      ))
    )
  
  # Return the dataframe
  return(the_df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
durham_2020 <- scrape.it(
  PWSID = "03-32-010",
  year = 2020
)

durham_2020 %>% 
  ggplot(aes(x = Date, y = max_day_use)) +
  geom_line() +
  labs(
    x = "Month",
    y = "Maximum Daily Withdrawal (MGD)",
    title = "Durham Maximum Daily Water Withdrawal, 2020"
  )

glimpse(durham_2020)

```

8. Use the function above to extract data for Cary (PWSID = '03-32-010') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals. NOTE FROM GINA-The assignment lists Cary's PWSID as `03-32-010`. The NC DEQ Local Water Supply Planning website identifies Cary as PWSID `03-92-020`, so `03-92-020`, so that is what was used for Cary data

```{r fetch.and.plot.Cary.2020.data}
#8 
cary_2020 <- scrape.it(
  PWSID = "03-92-020",
  year = 2020
)

combined_2020 <- bind_rows(
  durham_2020,
  cary_2020
)

combined_2020 %>% 
  ggplot(aes(x = Date, y = max_day_use, 
             linetype = system_name)) +
  geom_line() +
  labs(
    x = "Month",
    y = "Maximum Daily Withdrawal (MGD)",
    linetype = "Water System",
    title = "Maximum Daily Water Withdrawals, 2020"
  )

```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
the_pwsids <- rep("03-92-020", 6)
the_years <- 2018:2023

cary_multiyear <- map2(
  the_pwsids,
  the_years,
  scrape.it
) %>% 
  bind_rows()

cary_multiyear %>% 
  ggplot(aes(x = Date, y = max_day_use)) +
  geom_line() +
  geom_smooth(method = "loess") +
  labs(
    x = "Month",
    y = "Maximum Daily Withdrawal (MGD)",
    title = "Cary Maximum Daily Water Withdrawal, 2018-2023"
  )

```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:
>yes. Cary's water usage appears to have a slight increasing trend over time. The loess line gradually increases from approximately 22 MGD in 2018 to approximately 24 MGD by 2022–2023. 